---
title: "Change Analysis"
output:
  pdf_document: default
  html_document: default
---

Another experiment to look at rate of change. Idea here is to compare the reported change against the cumulative sum. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(readr)

```

```{r, echo=FALSE}
us_states_raw <- read_csv("~/projects/covid-19-data/us-states.csv",
                      col_types = cols(date = col_date(format = "%Y-%m-%d")))

#us_states$state <- as.factor(as.factor(us_states$state))
#us_states <- mutate_if(us_states, is.character, as.factor)

df_us_states <- us_states_raw %>%
  mutate_if(is.character, as.factor) 
  
# prep variable scale for Y axis
max_y <- ceiling(max(log2(df_us_states['cases'])))
powers_of_2 = 2 ** (1:max_y)

```

## Extract daily changes
```{r, echo=FALSE}
# compute lag so that we can see actual change day by day
df_us_states <- df_us_states %>%
  group_by(state) %>%
  mutate(delta_cases =  cases - lag(cases, order_by=date), 
         delta_deaths = deaths - lag(deaths,order_by=date))

```

### Daily Cases

```{r, fig.width=12, fig.height=10, echo=FALSE, message=FALSE}
p <- ggplot(subset(df_us_states, !is.na(delta_cases)), aes(x=date, y=delta_cases)) +
  geom_col() + 
  facet_wrap( . ~ state) +
  ggtitle("Covid change in cases by state - linear") +
  theme(strip.text.x = element_text(size = 9)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 print(p)

```

### Daily Deaths

```{r, fig.width=12, fig.height=10, echo=FALSE, message=FALSE}
p <- ggplot(subset(df_us_states, !is.na(delta_deaths)), aes(x=date, y=delta_deaths)) +
  geom_col() + 
  facet_wrap( . ~ state) +
  ggtitle("Covid change in deaths by state - linear") +
  theme(strip.text.x = element_text(size = 9)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 print(p)

```

## Change over Cumulative Sum

### Change in Cases over Cumulative Sum

Use the log of the cumulative sum to better scale the values.

```{r, fig.width=12, fig.height=10, echo=FALSE, message=FALSE, warning=FALSE}
p <- ggplot(df_us_states, aes(x=cases, y=delta_cases)) +
  geom_line() + 
  facet_wrap( . ~ state) +
  scale_x_continuous(trans='log10') +
  ggtitle("Covid cases rate of change by state - linear/log-x") +
  #theme(axis.text.y = element_text(size = 6)) +
  theme(strip.text.x = element_text(size = 9)) +
  theme(strip.text.y = element_text(size = 9)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
print(p)

```
## Log of Change Over Cumulative Sum

Linear values are skewed by higher-magnitude values, so use a log(10) on each axis.
this better fits the exponential nature of the data anyway.

### Log/Log of Cases Over Cumulative Sum

```{r, fig.width=12, fig.height=10, echo=FALSE, message=FALSE, warning=FALSE}
#filtered_cases <- subset(df_us_states, cases > 0)
pq <- ggplot(df_us_states, aes(x=cases, y=delta_cases)) +
  geom_line() + 
  facet_wrap( . ~ state) +
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') +
  ggtitle("Covid cases rate of change by state - log/log") +
  #theme(axis.text.y = element_text(size = 6)) +
  theme(strip.text.x = element_text(size = 9)) +
  theme(strip.text.y = element_text(size = 9)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
print(pq)

```

### Log/Log of Deaths over Cumulative Sum

```{r, fig.width=12, fig.height=10, echo=FALSE, message=FALSE, warning=FALSE}
#filtered_cases <- subset(df_us_states, cases > 0)
pq <- ggplot(df_us_states, aes(x=deaths, y=delta_deaths)) +
  geom_line() + 
  facet_wrap( . ~ state) +
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') +
  ggtitle("Covid deaths rate of change by state") +
  #theme(axis.text.y = element_text(size = 6)) +
  theme(strip.text.x = element_text(size = 9)) +
  theme(strip.text.y = element_text(size = 9)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
print(pq)

```

## Weekly Rate of Changes

In order to smooth out the curves in the previous graphs, look at them on a weekly basis.

```{r, echo=FALSE}
# let's group weekly to aggregate the lags we've already conmputed
df_us_states_weekly <- df_us_states %>%
  group_by(state, week=week(date)) %>%
  summarize(weekly_delta_cases = sum(delta_cases),
            weekly_delta_deaths = sum(delta_deaths),
            weekly_cases = max(cases),
            weekly_deaths = max(deaths)) %>%
  ungroup() %>%
  arrange(state, week)
```

### Weekly Range of Change of Cases by State

```{r, fig.width=12, fig.height=10, echo=FALSE, message=FALSE, warning=FALSE}
#filtered_cases <- subset(df_us_states, cases > 0)
pq <- ggplot(df_us_states_weekly, aes(x=weekly_cases, y=weekly_delta_cases)) +
  geom_line() + 
  facet_wrap( . ~ state) +
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') +
  ggtitle("Covid - weekly cases rate of change by state") +
  #theme(axis.text.y = element_text(size = 6)) +
  theme(strip.text.x = element_text(size = 9)) +
  theme(strip.text.y = element_text(size = 9)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
print(pq)

```

### Weekly Range of Change of Deaths by State

```{r, fig.width=12, fig.height=10, echo=FALSE, message=FALSE, warning=FALSE}
pq <- ggplot(df_us_states_weekly, aes(x=weekly_deaths, y=weekly_delta_deaths)) +
  geom_line() + 
  facet_wrap( . ~ state) +
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') +
  ggtitle("Covid - weekly deaths rate of change by state") +
  #theme(axis.text.y = element_text(size = 6)) +
  theme(strip.text.x = element_text(size = 9)) +
  theme(strip.text.y = element_text(size = 9)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
print(pq)

```